Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-09 main.tex
Reorganize paragraphs

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer sience. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on its own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in particular\cite{}\todo{find references}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

@@

@@

@@

We consider Gentzen's sequent calculus system $\LK$ \cite{gentzen:35:I}.

\begin{figure}[h]
\begin{center}
\begin{prooftree}
  \infer0[$\ax$]{\vdash A, \neg A}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, \neg A}
  \infer2[$\cut$]{\vdash \Gamma, \Delta}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, B}
  \infer2[$\cand$]{\vdash \Gamma, \Delta, A \cand B}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A, B}
  \infer1[$\cor$]{\vdash \Gamma, A \cor B}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A, A}
  \infer1[$\ctr$]{\vdash \Gamma, A}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma}
  \infer1[$\wk$]{\vdash \Gamma, A}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
\infer0[$\ttt$]{\vdash \ttt}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma}
	\infer1[$\fff$]{\vdash \Gamma, \fff}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \infer1[$\forall$]{\vdash \Gamma, \forall x A}
\end{prooftree}
($x \notin fv(\Gamma)$)
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A[t/x]}
  \infer1[$\exists$]{\vdash \Gamma, \exists x A}
\end{prooftree}
\caption{Gentzen's sequent calculus $\LK$}
\end{center}
\end{figure}

@@

\subsection{First-order combinatorial proofs}
A cograph is \bfit{logical} if every vertex is labelled either by an atom or variable,
and it has at least one atom-labelled vertex. We denote by $label_G(v)$, or simply
$label(v)$ the label of a vertex in a logical cograph $G$. An atom-labelled vertex is called
a \bfit{literal} and a variable-labelled vertex is called a \bfit{binder}. A
binder labelled with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest strong module containing at least two vertices, including
the binder itself. An \bfit{$x$-literal} is a literal whose atom contains the variable $x$. An $x$-binder \bfit{binds} every $x$-literal in its scope.
In a logical cograph $G$, a binder $b$ is \bfit{existential} (resp. \bfit{universal}) if, for every other
vertex $v$ in its scope, we have $bv \in E_G$ (resp. $bv \notin E_G$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder and at least one
literal.

For a formula $A$, we can define its associated fograph $\Gr(A)$ inductively by: $\Gr(\alpha) = \cdot \alpha$ (a single-vertex graph whose vertex is labelled by $\alpha$); $\Gr(A \cor B) = \Gr(A) + \Gr(B)$ and $\Gr(A \cand B) = \Gr(A) \times \Gr(B)$.
The \bfit{binding graph} of a forgraph $G$ is the directed graph $(V_G, \{(b,
l) | b \text{ binds } l\})$.

>>>>

First-order predicate logic is one of the cornerstones of modern
logic. Since its formalisation by Frege~\cite{frege:79} it has seen a
growing usage in many fields of mathematics and computer sience. Since
the development of proof theory by Hilbert~\cite{hilbert:22},
\emph{proofs} became first-class citizens as mathematical objects that
could be studied on its own. Since Gentzen's \emph{sequent
calculus}~\cite{gentzen:35:I,gentzen:35:II}, many other proof systems
have been developed that allow the implementation of efficient proof
search, for example \emph{analytic tableaux}~\cite{smullyan:68} or
\emph{resolution}~\cite{robinson:65}. Despite the immense progress
that has been made in proof theory in general and in the area of
automated and interactive theorem provers in particular\cite{paulson:isabelle}\todo{find references}, we still have
no satisfactory notion of proof identity for first-order logic. In
this respect, proof theory is quite different from any other
mathematical field. For example in group theory, two groups are
\emph{the same} iff they are isomorphic; in topology, two spaces are
\emph{the same} iff they are homeomorphic; etc. In proof theory, we
have no such notion telling us when two proofs are \emph{the same},
even though Hilbert was considering this problem as possible 24th
problem for his famous lecture~\cite{hilbert:00} in
1900~\cite{thiele:03}, before proof theory existed as a mathematical
field.

@@

We consider Gentzen's sequent calculus system $\FOLK$ \cite{gentzen:35:I}.

\begin{figure}[h]
\begin{center}
\begin{prooftree}
  \infer0[$\ax$]{\vdash A, \neg A}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, \neg A}
  \infer2[$\cut$]{\vdash \Gamma, \Delta}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \hypo{\vdash \Delta, B}
  \infer2[$\cand$]{\vdash \Gamma, \Delta, A \cand B}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A, B}
  \infer1[$\cor$]{\vdash \Gamma, A \cor B}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A, A}
  \infer1[$\ctr$]{\vdash \Gamma, A}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma}
  \infer1[$\wk$]{\vdash \Gamma, A}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
\infer0[$\ttt$]{\vdash \ttt}
\end{prooftree}
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma}
	\infer1[$\fff$]{\vdash \Gamma, \fff}
\end{prooftree}
\\[1.5ex]
\begin{prooftree}
  \hypo{\vdash \Gamma, A}
  \infer1[$\forall$]{\vdash \Gamma, \forall x A}
\end{prooftree}
($x \notin fv(\Gamma)$)
\qquad
\begin{prooftree}
  \hypo{\vdash \Gamma, A[t/x]}
  \infer1[$\exists$]{\vdash \Gamma, \exists x A}
\end{prooftree}
\caption{Gentzen's sequent calculus $\FOLK$}
\end{center}
\end{figure}

@@

A cograph is \bfit{logical} if every vertex is labelled either by an atom or variable,
and it has at least one atom-labelled vertex. We denote by $label_G(v)$, or simply
$label(v)$ the label of a vertex in a logical cograph $G$. An atom-labelled vertex is called
a \bfit{literal} and a variable-labelled vertex is called a \bfit{binder}. A
binder labelled with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest strong module containing at least two vertices, including
the binder itself. An \bfit{$x$-literal} is a literal whose atom contains the variable $x$. An $x$-binder \bfit{binds} every $x$-literal in its scope.
In a logical cograph $G$, a binder $b$ is \bfit{existential} (resp. \bfit{universal}) if, for every other
vertex $v$ in its scope, we have $bv \in E_G$ (resp. $bv \notin E_G$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder and at least one
literal.

For a formula $A$, we can define its associated fograph $\Gr(A)$ inductively by: $\Gr(\alpha) = \cdot \alpha$ (a single-vertex graph whose vertex is labelled by $\alpha$); $\Gr(A \cor B) = \Gr(A) + \Gr(B)$ and $\Gr(A \cand B) = \Gr(A) \times \Gr(B)$.
The \bfit{binding graph} of a forgraph $G$ is the directed graph $(V_G, \{(b,
l) \ | \ b \text{ binds } l\})$.

@@

\subsection{First-order combinatorial proofs}

@@

@@

\subsection{First-order combinatorial proofs}
